---
title: "Homework 2 on Newton's methods"
author: "Leave your name and uni here"
date: "Due: 03/18/2020, Wednesday, by 1pm"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(modelr)
library(viridis)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```

# Problem 1
Design an optimization algorithm to find the minimum of the continuously differentiable function $f(x) =-e^{-1}\sin(x)$ on the closed interval $[0,1.5]$. Write out your algorithm and implement it into \textbf{R}.


# Answer: your answer starts here...

```{r }
#R codes:

f = function(x){
  return(-sin(x)*exp(-x))
}


optimize(f, interval = c(0, 1.5))
```

```{r}
curve(f, from = 0, to = 1.5)
```


```{r}
w = .618
a =  0
b = 1.5
tol = 1e-6
i = 0  # iteration index
x1 = (1 - w) * (b - a) + a
x2 = x1 + w * (b - a) * (1 - w)
res = c(a, b, x1, x2)

findmin = function() {
  while (abs(b - a) > tol) {
    i <- i + 1
    if (f(b) > f(a)) {
      b <- x2
      x1 <- (1 - w) * (b - a) + a
      x2 <- x1 + w * (b - a) * (1 - w)
    }
    else {
      a <- x1
      x1 <- x2
      x2 <- x1 + w * (b - a) * (1 - w)
    }
    res <- rbind(res, c(a, b, x1, x2))
  }

    return(res)
}

min = last(findmin())
```

The minimum according to the golden ratio search method is `r round(min, digits = 3)`. This closely approximates the result of optimize(), `r optimize(f, interval = c(0, 1.5))$minimum %>% round(digits =3)`.

# Problem 2
The Poisson distribution is often used to model ``count'' data ---
e.g., the number of events in a given time period.  
The Poisson regression model states that
$$Y_i \sim \textrm{Poisson}(\lambda_i),$$
where
$$\log \lambda_i = \alpha + \beta x_i $$
 for some explanatory variable
$x_i$.  The question is how to estimate $\alpha$ and $\beta$ given a
set of independent data $(x_1, Y_1), (x_2, Y_2), \ldots, (x_n, Y_n)$.
\begin{enumerate}
\item Modify the Newton-Raphson function from the class notes to include
a step-halving step.
\item Further modify this function to ensure that the direction of the
step is an ascent direction.   (If it is not, the program should take appropriate
action.)
\item Write code to apply the resulting modified Newton-Raphson function
to compute maximum likelihood estimates for $\alpha$ and $\beta$
in the Poisson regression setting.
\end{enumerate}

\vskip 5mm
\noindent
The Poisson distribution is given by
$$P(Y=y) = \frac{\lambda^y e^{-\lambda}}{y!}$$
for $\lambda > 0$. 


# Answer: your answer starts here...

```{r }
#R codes:
```


\paragraph{problem 3}

\vskip 10pt
Consider the ABO blood type data, where you have $N_{\mbox{obs}} = (N_A,N_B,N_O,N_{AB}) = ( 26, 27, 42, 7)$.

\begin{itemize}
\item design an EM algorithm to estimate  the allele frequencies, $P_A$, $P_B$ and $P_O$; and 

\item Implement your algorithms in R, and present your results..
\end{itemize}

# Answer: your answer starts here...

```{r }
#R codes:
```

